---
title: "Effects of weather events (in the U.S.) on population and economy."
author: "Daniel Carhuas"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: 
 html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 8, fig.width = 12)
```

## Synopsis

In this report we explore the U.S. National Oceanic and Atmospheric Administration (NOAA) storm database.
National Oceanic and Atmospheric Administration (NOAA). This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. We will analyze the data to answer the following questions :

* In the United States, what types of events (indicated by the variable EVTYPE) are most detrimental to the health of the population?
* In the United States, what types of events have the greatest economic consequences? 

Our analysis concludes that: 

* Tornadoes are the most damaging to the health of the population.
* Floods are the most damaging overall, while hurricanes/typhoons are the most devastating per event.

Translated with DeepL.com (free version)

 
## Data processing

Library loading
```{r}
if(!require(dplyr)) install.packages("ggplot2", dependencies=TRUE)
if(!require(ggplot2)) install.packages("ggplot2", dependencies=TRUE)
if(!require(scales)) install.packages("ggplot2", dependencies=TRUE)
if(!require(gridExtra)) install.packages("ggplot2", dependencies=TRUE)

library(dplyr)
library(ggplot2)
library(scales)
library(gridExtra)
```

Downloading and reading the file

```{r cache=TRUE}

file_storm <- "./repdata_data_StormData.csv.bz2"
if(!file.exists(file_storm)) {
    url_storm <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(url = url_storm, destfile = file_storm)
}

df_storm <- read.csv(bzfile(file_storm))
```
After loading the data, we will check the first rows and the data structure.
```{r}
head(df_storm)
```
For the purpose of our analysis, we will select only the columns we are interested in. 
    1. FATALITIES AND INJURIES columns inform us about the effect of storms and meteorological phenomena on the health of the population 2. 
    2. The columns PROPDMG, PROPDMGEXP and CROPDMG, CROPDMGEXP tell us about the economic damage caused by storms and weather events by estimating the approximate losses (in dollars) of goods and crops.
    
```{r}
colsToSelect <- c("BGN_DATE","EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")

df_storm_short <- df_storm[, colsToSelect]
head(df_storm_short)

```

Checking unique values in PROPDMGEXP and CROPDMGEXP to understand the different types of characters used to indicate the multiplication factor to be applied to PROPDMG and CROPDMG respectively.

```{r}
unique(df_storm_short$PROPDMGEXP)
unique(df_storm_short$CROPDMGEXP)
```
From the above we can understand the following correspondences: 

* h/H = cien /'00
* k/K = mil / '000 
* m/M = millón / '000.000
* b/B = mil millones / '000.000.000

Add a column for transforming character symbols into their corresponding
multiplication factor
```{r}

df_storm_short <- df_storm_short %>% 
    mutate(PROPDMGEXP2 = case_when(
        (PROPDMGEXP == "h" | PROPDMGEXP == "H") ~ 100,
        (PROPDMGEXP == "k" | PROPDMGEXP == "K") ~ 1000,
        (PROPDMGEXP == "m" | PROPDMGEXP == "M") ~ 1000000,
        (PROPDMGEXP == "b" | PROPDMGEXP == "B") ~ 1000000000,
        PROPDMGEXP == "" ~ 1
    ),
    CROPDMGEXP2 = case_when(
        (CROPDMGEXP == "h" | CROPDMGEXP == "H") ~ 100,
        (CROPDMGEXP == "k" | CROPDMGEXP == "K") ~ 1000,
        (CROPDMGEXP == "m" | CROPDMGEXP == "M") ~ 1000000,
        (CROPDMGEXP == "b" | CROPDMGEXP == "B") ~ 1000000000,
        CROPDMGEXP == "" ~ 1
    ))
       
```
For the rest of the characters, we are NOT able to interpret the meaning.
meaning. We will check how much data there is, apart from the characters we have identified, to understand the impact these data will have on our analysis

```{r}
table(df_storm_short$PROPDMGEXP2, useNA = "ifany")
table(df_storm_short$CROPDMGEXP2, useNA = "ifany")
```
We can see here that the NA values, or the # of observations for which we were unable to interpret the characters, represent a very small percentage percent of the total number of observations. We will ignore the transformation of these observations.

calculation of the economic damage value with the derived multiplication factor derived factor
```{r}

df_storm_short <- df_storm_short %>% 
    mutate(PROPDMGVAL = PROPDMG*PROPDMGEXP2,
           CROPDMGVAL = CROPDMG*CROPDMGEXP2)
```
## Results

#### Question 1: In the United States, what types of events (indicated by the variable EVTYPE) are most damaging to the health of the population?
```{r}

#fatalities2 <- tapply(df_storm_short$FATALITIES, df_storm_short$EVTYPE, sum)
#fatalities2 <- sort(fatalities2, decreasing = TRUE)
#fatalities2 <- fatalities2[1:10]

fatalities <- aggregate(FATALITIES ~ EVTYPE, df_storm_short, sum)
fatalities <- fatalities[order(-fatalities$FATALITIES), ]
fatalities <- fatalities[1:10,]
fatalities$EVTYPE <- factor(fatalities$EVTYPE, levels = fatalities$EVTYPE)

p1 <- ggplot(fatalities, aes(x = EVTYPE, y = FATALITIES)) + geom_bar(stat = "identity") + xlab("Event Type") + ylab("No of Fatalities") + ggtitle("Most Fatalities per Weather Event") + theme(axis.text.x = element_text(angle = 45, hjust = 1))

#injuries2 <- tapply(df_storm_short$INJURIES, df_storm_short$EVTYPE, sum)
#injuries2 <- sort(injuries2, decreasing = TRUE)
#injuries2 <- injuries2[1:10]

injuries <- aggregate(INJURIES ~ EVTYPE, df_storm_short, sum)
injuries <- injuries[order(-injuries$INJURIES),]
injuries <- injuries[1:10,]
injuries$EVTYPE <- factor(injuries$EVTYPE, levels = injuries$EVTYPE)

p2 <- ggplot(injuries, aes(x = EVTYPE, y = INJURIES)) + geom_bar(stat = "identity") + xlab("Event Type")+ ylab("No of Injuries") + ggtitle("Most Injuries per Weather Event") + theme(axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(p1,p2, ncol = 2)

```
#### Answer 1: Tornado is the most damaging weather event to the health of the population.

#### Question 2: In the United States, what types of events have the greatest economic consequences?
```{r}
damages <- aggregate(PROPDMGVAL + CROPDMGVAL ~ EVTYPE, df_storm_short, sum)
names(damages) <- c("EVTYPE", "TOTDMG")
damages <- damages[order(-damages$TOTDMG), ]
damages <- damages[1:10, ]
damages$EVTYPE <- factor(damages$EVTYPE, levels = damages$EVTYPE)

p3 <- ggplot(damages, aes(x = EVTYPE, y = TOTDMG)) + geom_bar(stat = "identity") + xlab("Event Type") +  ylab("Damages $") + ggtitle("Most Damgages (in USD) per Weather Event") + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7)) + scale_y_continuous(labels = label_number(suffix = " Bn", scale = 1e-09))

damages_nObs <- df_storm_short %>% 
    count(EVTYPE)
damages_nObs <- damages_nObs[damages_nObs$EVTYPE %in% damages[,1],]
damages_nObs$EVTYPE <- factor(damages_nObs$EVTYPE, levels = damages$EVTYPE)

p4<- ggplot(damages_nObs, aes(x = EVTYPE, y = n)) + geom_bar(stat = "identity") + geom_text(aes(label = n), vjust = -0.3, size = 3) + xlab("Event Type") + ylab("No of Events") + ggtitle("No of Events per Event Type") + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7)) + scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-03))

grid.arrange(p3,p4)
```
#### Answer 2: Floods have had the greatest economic consequences. However, if you look at the corresponding number of events, hurricanes and typhoons are the deadliest weather events with the greatest economic damage per event.